---
title: "PopGen_Chapter2.rmarkdown"
author: "Nicole, Parisa, Sol"
date: "2024-09-10"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages(blogdown)
library(blogdown)
library(ggplot2)
```

## Problem 2.1 - Sol

In a population of size $N$, there are $2N$ alleles for a given locus (because each diploid individual has 2 copies of each allele). Therefore the chance that a given allele, $A_1$, is chosen at random is $\frac{1}{2N}$ and the chance that it is not chosen is $1-\frac{1}{2N}$. 

Therefore the chance that $A_1$ will not be chosen in 2N rounds with replacement (based on the simulation shown in Fig 2.1 in the book) is $(1-\frac{1}{2N})^{2N}$ (ie: the chance it will not be chosen each round to the power of the number of rounds in the simulation). 

Thus, the probability that it *will* be chosen at least 1x (ie the probability that a particular allele has at least one copy in the next generation) = $1-(1-\frac{1}{2N})^{2N}$

As the book mentions, this becomes independent of $N$ as the population grows very large, therefore we can calculate the probability for any population with an extremely large N based on the following equation from the book:

$(A.7)$ $lim_{\epsilon \rightarrow0}(1+\epsilon x)^{1/\epsilon} = e^x$

We know that as $N$ increases the value of $\frac{1}{2N}$ approaches 0. Thus we can use equation A.7 above with $\epsilon=\frac{1}{2N}$:

$(1-\frac{1}{2N})^{2N} \rightarrow lim_{\frac{1}{2N} \rightarrow0}(1+(\frac{1}{2N}x))^{2N}$ 

for $(1-\frac{1}{2N})^{2N}$ to equal $(1+(\frac{x}{2N}))^{2N}$, x must equal -1.

Therefore, based on equation A.7 where $\epsilon=\frac{1}{2N}$: and $x=-1$:

$lim_{\frac{1}{2N} \rightarrow0}(1-\frac{1}{2N})^{2N} = e^{-1} = 0.37$

Thus the chance the $A_1$ is never chosen (ie does not appear in the next generation in 0.37). 1-0.37 = 0.63. Therefore in any extremely large population where $\frac{1}{2N}\rightarrow0$ the probability that at least one copy of an allele will appear in the next generation is 0.63.

## Problem 2.2 -- Parisa

```{r problem2.2}
library(ggplot2)
library(reshape2)



twoN <- 40
p <- 0.2
num_generations <- 100
num_populations <- 5

freq_A1 <- matrix(0, nrow = num_generations, ncol = num_populations)

freq_A1[1, ] <- p

for (pop in 1:num_populations) {
  for (generation in 2:num_generations) {
    if (is.na(freq_A1[generation - 1, pop]) || freq_A1[generation - 1, pop] == 0 || freq_A1[generation - 1, pop] == 1) {
      freq_A1[generation, pop] <- NA  
    } else {
      num_A1 <- sum(rbinom(twoN, 1, freq_A1[generation - 1, pop]))
      freq_A1[generation, pop] <- num_A1 / twoN
    }
  }
}
freq_df <- as.data.frame(freq_A1)
freq_df$Generation <- 1:num_generations
freq_melted <- melt(freq_df, id.vars = "Generation", variable.name = "Population", value.name = "Frequency")

p <- ggplot(freq_melted, aes(x = Generation, y = Frequency, color = Population, group = Population)) +
  geom_line(size = 1.2) +
  labs(title = "Genetic Drift: Frequency of A1 over Generations",
       x = "Generation", y = "Frequency of A1") +
  theme_minimal() +
  theme(legend.position = "right",
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.justification = c(1, 0.5))

p + theme(legend.position = "top", 
          legend.direction = "horizontal",
          legend.box = "horizontal") +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))
```

## Problem 2.3 

```{r problem2.3}

```

## Problem 2.4 - Parisa

```{r problem2.4}
H0 <- 1
t <- 0:100 
N_values <- c(1, 10, 100, 1000000)

results <- data.frame(Generation = rep(t, times = length(N_values)),
                      H = NA,
                      G = NA,
                      N = rep(N_values, each = length(t)))

for (N in N_values) {
  H <- H0 * (1 - 1/(2*N))^t
  G <- 1 - H
  
  results$H[results$N == N] <- H
  results$G[results$N == N] <- G
}

p1 <- ggplot(results, aes(x = Generation, y = H, color = as.factor(N))) +
  geom_line() +
  labs(title = "Heterozygosity over Time",
       y = "H(t)",
       x = "Generation",
       color = "N (Population Size)") +
  theme_minimal()
print(p1)
p2 <- ggplot(results, aes(x = Generation, y = G, color = as.factor(N))) +
  geom_line() +
  labs(title = "Fixation Probability over Time",
       y = "G(t)",
       x = "Generation",
       color = "N (Population Size)") +
  theme_minimal()
print(p2)
```

## Problem 2.5 

```{r problem2.5}
#graph formula 2.4 and 2.5 for N 1 to 100
library(data.table)
library(ggplot2)

#create df for simulation
df = data.table(N = 1:100)

#loop through all N to see if the apprx estimate is close to the exact
for (n in 1:100){
  exact = (-log(2)) / (log(1-1/(2*n)))
  apprx = 2*n*log(2)
  df[n, 'Exact' := exact]
  df[n, 'Apprx' := apprx]
}

#view df
#head(df)

#melt df to plot all N
df_melted <- melt(df, id.vars = "N", variable.name = "Sample", value.name = "Rate")

#plot results
ggplot(df_melted, aes(x = N, y = Rate, color = Sample)) +
  geom_line() +
  theme_classic()+
  scale_colour_manual("", values = c("black",'#b2182b'))+
  labs(title = "Reduce Ht by Half Simulation", x = "N", y = "Rate",
       color = "Iterations")


```

## Problem 2.6 - Sol
*Because calligraphic letters don't render correctly into html I will be using $G'$ to represent the calligraphic G in the book.*

Remember that:  
G = Homozygosity of a population   
$G'$ = Probability that 2 alleles that differ by origin are identical by state

Based on the problem, we will define the homozygosity of a population (G) as the probability that two alleles chosen at random from the population *with replacement* are identical by state.

First, we want to show that this definition is equivalent to the definition given in Equation 1.3:

$(1.3)$ G=$\sum^{k}_{i=1}p^2_i$

*where $k$ = \#alleles and $p_i$ = frequency of allele $A_i$*

The probability of 2 of the same allele getting chosen for any allele, $A_i$, is $p_i*p_i=p_i^2$  
(ie the probability of getting $A_i$ * the probability of getting $A_i$  -- since the events are independent)

Summed for all alleles = $\sum^{k}_{i=1}p^2_i$ (which is equation 1.3)

Next, we want to show that:

G$=\frac{1}{2N}+(1-\frac{1}{2N})G'$

Remember that $2N$ is the number of alleles at a given locus in a population (2 * the number of people).

Thus, $\frac{1}{2N}$ is the chance of a specific allele (by origin) being chosen. 

For a given allele, $A_i$, the chance of the same allele by origin being chosen twice (since we are sampling with replacement) = $\frac{1}{2N}$. 

Therefore, the probability of getting two alleles that differ by origin = $1-\frac{1}{2N}$.

As previously defined, $G'$ = the probability of 2 alleles that differ by origin being identical by state. 

Therefore the equation given above (G$=\frac{1}{2N}+(1-\frac{1}{2N})G'$) breaks down to:

*G = (the probability of getting the same allele by origin 2x) + (the probability of getting 2 alleles that are different by origin x the probability of 2 alleles that differ by origin being identical by state)*

This makes sense because there are two ways for two alleles to be identical by state in this scenario: (1) They are identical by origin and (2) they are different by origin but identical by state.

This concept is demonstrated in Fig 2.3 in the book. 

